---
title: "Course Project 1"
author: "David Graber"
date: "August 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading in the Data
```{r readdata}
activity <- read.csv(file = "activity.csv", header = TRUE)
```

## Histogram of Daily Steps Taken
```{r histogram}
daily <- tapply(activity$steps, activity$date, FUN = sum)
hist(daily, main = "Daily Steps Taken", xlab = "Total Daily Steps")
```

## Mean and Median Steps Taken:
```{r}
dailymean <- mean(daily, na.rm = TRUE)
dailymed <- median(daily, na.rm = TRUE)
```

The mean number of steps per day is `r dailymean`.  
The median number of steps per day is `r dailymed`.

## Time Series of Daily Activity
```{r dailytimeseries}
timeseries <- tapply(activity$steps, activity$interval, FUN = mean, na.rm = TRUE)
plot(timeseries, type = "l", ylab = "Average Steps", xlab = "Time Interval (Minutes)")
maxinterval <- activity$interval[which.max(timeseries)]
```
On average, the `r maxinterval` minute time interval contains the maximum number of steps.

## Missing Values
```{r}
missing <- sum(is.na(activity$steps))
```
There are `r missing` NA values in the dataset.

The "NA" values will be replaced with the mean value for their time interval.

```{r replaceNA}
fixed <- activity
toreplace <- which(is.na(activity$steps))
fixed$steps[toreplace] <- timeseries[which(names(timeseries) == fixed$interval[toreplace])]
```
Replicating the earlier calculations on the "fixed" dataset:

```{r histogram2}
daily2 <- tapply(fixed$steps, fixed$date, FUN = sum)
hist(daily2, main = "Daily Steps Taken", xlab = "Total Daily Steps")
```
```{r}
dailymean2 <- mean(daily2, na.rm = TRUE)
dailymed2 <- median(daily2, na.rm = TRUE)
```

The mean number of steps per day was `r dailymean` and is now `r dailymean2`.  
The median number of steps per day was `r dailymed` and is now `r dailymed2`.  
Adding mean values to replace NAs did not affect the overall averages.


## Comparing Weekdays and Weekends
Add new factor column indicating weekdays or weekends, then take averages and plot
```{r weekday}
fixed$weekday <- ifelse(weekdays(as.Date(fixed$date)) %in% c("Saturday", "Sunday"), "weekend", "weekday")

averages <- as.data.frame(tapply(fixed$steps, list(fixed$interval, fixed$weekday), FUN = mean, na.rm = TRUE))

library(data.table)
setDT(averages, keep.rownames = TRUE)[]
names(averages)[1] <- "interval"

par(mfrow = c(2,1))
with(averages, plot(interval, weekday, main = "Average Steps by Time Interval on Weekdays", type = "l", xlab = "5-min Time Interval", ylab = "Mean Steps Taken", ylim = c(0,250), xaxt = "n"))
axis(1, xaxp = c(0,2400, 20), las = 2)
with(averages, plot(interval, weekend, main = "Average Steps by Time Interval on Weekends", type = "l", xlab = "5-min Time Interval", ylab = "Mean Steps Taken", ylim = c(0,250), xaxt = "n"))
axis(1, xaxp = c(0,2400, 20), las = 2)
```
